Fast and scalable Gaussian process modeling with applications to astronomical time series \vspace{-3\baselineskip}

### 2017-03-15 paper/ms.tex
benchmark words [ci skip]

The extended system defined in the previous section is sparse with fewer than
a few percent non-zero entries and band structure.
In this \sectionname, we empirically investigate the performance and scaling
of three different algorithms for solving this extended systems:

@@

\item A general sparse LU solver that exploits the sparsity but not the band
    structure \citep{Guennebaud:2010}.

\item A simple algorithm for computing the LU decomposition for banded
    matrices using Gaussian elimination \citep{Press:1992, Press:2007},

\item The general banded LU decomposition implementation from
    LAPACK\footnote{We use the \texttt{dgbtrf} and \texttt{dgbtrs} methods
        from LAPACK.} \citep{Anderson:1999} using optimized BLAS routines, and

@@

\citet{Ambikasaran:2015} found an empirical scaling of $\mathcal{O}(N\,J^2)$
with their method that used the sparse LU decomposition implemented in the
\project{SuperLU} package \citep{Demmel:1999}.
The theoretical scaling for a band LU decomposition is $\mathcal{O}(N\,J^3)$
because the dimension of the extended matrix scales as $N\,J$ and the
bandwidth scales with $J$ \citep{Press:1992, Press:2007}.
We find that, while the vanilla banded solver scales as expected, the LAPACK
algorithm with tuned BLAS routines scales empirically as $\mathcal{O}(N\,J^2)$
empirically and offers the fastest solves for $J \gtrsim 8$ on all platforms
that we tested.

>>>>

The extended system defined in the previous section is sparse with fewer than
a few percent non-zero entries and band structure.
In this \sectionname, we empirically investigate the performance and scaling
of three different algorithms for solving this extended system:

@@

\item \texttt{vanilla}: A simple algorithm for computing the LU decomposition
    for banded matrices using Gaussian elimination \citep{Press:1992,
    Press:2007},

\item \texttt{lapack}: The general banded LU decomposition implementation from
    LAPACK\footnote{We use the \texttt{dgbtrf} and \texttt{dgbtrs} methods
    from LAPACK.} \citep{Anderson:1999} using optimized BLAS
    routines\footnote{The experiments below use the Intel Math Kernel Library
    (\project{MKL}): \url{https://software.intel.com/en-us/intel-mkl}.}, and

\item \texttt{sparse}: A general sparse LU solver~--~the \texttt{SparseLU}
    solver from \project{Eigen} \citep{Guennebaud:2010}~--~that exploits the
    sparsity but not the band structure.

@@

\citet{Ambikasaran:2015} found an empirical scaling of $\mathcal{O}(N\,J^2)$
with their method that used the sparse LU decomposition implemented in the
\project{SuperLU} package \citep{Demmel:1999}.
The theoretical scaling for a band LU decomposition is $\mathcal{O}(N\,J^3)$
because the dimension of the extended matrix scales as $N\,J$ and the
bandwidth scales with $J$ \citep{Press:1992, Press:2007}.
We find that, while the \texttt{vanilla} solver scales as expected, the
\texttt{lapack} implementation scales empirically as $\mathcal{O}(N\,J^2)$ and
offers the fastest solves for $J \gtrsim 8$ on all platforms that we tested.

The benchmark experiments shown here were performed on a MacBook Pro with two
2.6~GHz CPUs but we find similar results on a Dell workstation with 16 2.7~GHz
CPUs and running Ubuntu.
